Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-11 TOMS_Paper/assembly.tex
Some other minor fixes

Our research has resulted in the development of a compiler, COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, fully integrated with the Firedrake framework~\cite{Luporini}. Besides separating the mathematical domain, captured by the form compiler at an higher level of abstraction, from the optimization process, COFFEE also aims to be platform-agnostic. The code transformations occur on an intermediate representation of the assembly operation, which is ultimately translated into platform-specific code. Domain knowledge is exploited in two ways: for simplifying the implementation of code transformations and to make them extremely effective. Domain knowledge is conveyed to COFFEE from the higher level (the form compiler in the case of Firedrake, although any user-provided code would be acceptable) through suitable annotations attached to the input. For example, when the input is in the form of an abstract syntax tree produced by the form compiler, specific nodes are decorated so as to drive the optimization process. Although COFFEE has been thought of as a multi-platform optimizing compiler, our performance evaluation so far has been restricted to standard CPU platforms only. We emphasize once more, however, that the transformations applicable by both the Expression Rewriter (Section~\ref{sec:expr-rewriting}) and the Code Specializer (Section~\ref{sec:code-spec}) would work on generic accelerators as well.

@@

Expression rewriting and code specialization have been implemented in COFFEE, which in turn is integrated with the Firedrake framework. Therefore, to testify the goodness of our approach, we provide an extensive and unprecedented performance evaluation across a number of forms of increasing complexity, including some based on hyperelasticity. We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions. To clearly distinguish the improvement achieved by this work, we will compare, for each test case, four sets of code variants: 1) unoptimized code, i.e. a local assembly routine as returned from the form compiler; 2) code optimized by FEniCS, i.e. the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; code optimized by expression rewriting and code specialization as described in this paper. Notable performance improvements of 4) over 1), 2) and 3) are reported and discussed.

@@

\subsection{Avoiding Iteration on Zero-blocks by Symbolic Execution}
\label{sec:zeros}
We aim at skipping arithmetic operations over blocks of zero-valued entries in basis functions (and their derivatives) arrays. Zero-valued columns arise, for example, when taking derivatives on a reference element and when employing mixed elements. In~\cite{quadrature1}, a technique to avoid operations on zero-valued columns based on the use of indirection arrays was described (e.g. \texttt{A[B[i]]}, where \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in \texttt{A}) and implemented in the context of FEniCS. We will evaluate our approach and compare it to this pioneering work in Section~\ref{sec:perf-results}. Essentially, our strategy avoids indirection arrays in the generated code, which otherwise would break the optimizations applicable at the Code Specializer level, including SIMD vectorization.

@@

Table~\ref{table:autotuner-output} shows the number a given transformation has been selected by the autotuner. To not hinder readability, the output has been grouped by form, rather than showing the selected optimization strategy for each $[form, q, p, nf]$. Values in the \texttt{rewrite strategy} column can either be \texttt{base} or \texttt{aggressive}, as explained in Section~\ref{sec:autotune}: in the former case, only generalized loop-invariant code motion in applied; in the latter case, the rewrite rules are recursively applied as extensive as possible. If the value of the column is \texttt{aggressive} (\texttt{base}) it means that in the majority of cases the \texttt{aggressive} (\texttt{base}) strategy prevailed on the \texttt{base} (\texttt{aggressive}) one. Precomputation of integration-dependent terms was explained in Section~\ref{sec:expr-rewriter-obj}. The \texttt{split} column refers to the expression splitting transformation (~\cite{Luporini}, and summary in Section~\ref{sec:summary-opts}). The \texttt{unroll} column indicates the application of explicit unrolling. Other columns are of obvious meaning. Values in the various columns illustrate the number of problem instances out of the total (column \texttt{mumber of variants}) in which an optimization was activated. Obviously, more transformations are typically used in combination in a same problem.

>>>>

Our research has resulted in the development of a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Finit Element local assEmbly.}, fully integrated with the Firedrake framework~\cite{Luporini}. Besides separating the mathematical domain, captured by the form compiler at an higher level of abstraction, from the optimization process, COFFEE also aims to be platform-agnostic. The code transformations occur on an intermediate representation of the assembly operation, which is ultimately translated into platform-specific code. Domain knowledge is exploited in two ways: for simplifying the implementation of code transformations and to make them extremely effective. Domain knowledge is conveyed to COFFEE from the higher level (the form compiler in the case of Firedrake, although any user-provided code would be acceptable) through suitable annotations attached to the input. For example, when the input is in the form of an abstract syntax tree produced by the form compiler, specific nodes are decorated so as to drive the optimization process. Although COFFEE has been thought of as a multi-platform optimizing compiler, our performance evaluation so far has been restricted to standard CPU platforms only. We emphasize once more, however, that the transformations applicable by both the Expression Rewriter (Section~\ref{sec:expr-rewriting}) and the Code Specializer (Section~\ref{sec:code-spec}) would work on generic accelerators as well.

@@

Expression rewriting and code specialization have been implemented in COFFEE, which in turn is integrated with the Firedrake framework. Therefore, to testify the goodness of our approach, we provide an extensive and unprecedented performance evaluation across a number of forms of increasing complexity, including some based on complex hyperelasticity models. We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions. To clearly distinguish the improvement achieved by this work, we will compare, for each test case, four sets of code variants: 1) unoptimized code, i.e. a local assembly routine as returned from the form compiler; 2) code optimized by FEniCS, i.e. the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; code optimized by expression rewriting and code specialization as described in this paper. Notable performance improvements of 4) over 1), 2) and 3) are reported and discussed.

@@

\subsection{Avoiding Iteration on Zero-valued Blocks by Symbolic Execution}
\label{sec:zeros}
We aim at skipping arithmetic operations over blocks of zero-valued entries in basis functions (and their derivatives) arrays. Zero-valued columns arise, for example, when taking derivatives on a reference element and when employing mixed elements. In~\cite{quadrature1}, a technique to avoid operations on zero-valued columns based on the use of indirection arrays was described (e.g. \texttt{A[B[i]]}, where \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in \texttt{A}) and implemented in the context of FEniCS. We will evaluate our approach and compare it to this pioneering work in Section~\ref{sec:perf-results}. Essentially, our strategy avoids indirection arrays in the generated code, which otherwise would break the optimizations applicable at the Code Specializer level, including SIMD vectorization.

@@

Table~\ref{table:autotuner-output} shows the number a given transformation has been selected by the autotuner. To not hinder readability, the output has been grouped by form, rather than showing the selected optimization strategy for each $[form, q, p, nf]$. Values in the \texttt{rewrite strategy} column can either be \texttt{base} or \texttt{aggressive}, as explained in Section~\ref{sec:autotune}: in the former case, only generalized loop-invariant code motion in applied; in the latter case, the rewrite rules are recursively applied as extensive as possible. If the value of the column is \texttt{aggressive} (\texttt{base}) it means that in the majority of cases the \texttt{aggressive} (\texttt{base}) strategy prevailed on the \texttt{base} (\texttt{aggressive}) one. Precomputation of integration-dependent terms was explained in Section~\ref{sec:expr-rewriter-obj}. The \texttt{split} column refers to the expression splitting transformation (\cite{Luporini}, and summary in Section~\ref{sec:summary-opts}). The \texttt{unroll} column indicates the application of explicit unrolling. Other columns are of obvious meaning. Values in the various columns illustrate the number of problem instances out of the total (column \texttt{mumber of variants}) in which an optimization was activated. Obviously, more transformations are typically used in combination in a same problem.
